---
title: "PA1_template.Rmd"
author: "Andy Majumdar"
date: "Sunday, February 19, 2017"
output: html_document
---

## 1: Code for reading in the dataset and/or processing the data

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [Download from Here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The absolute location of the local directory where the file is stored and name of the file is "actvity.csv" is given below. Please change after you download the file from the following location:


```{r}
fpath <- "C:\\Users\\Arindam\\Desktop\\RWk1\\Wk2\\activity.csv"
# load the csv in data frame

activity_all <- read.csv(fpath)
activity_all$steps <- as.numeric(activity_all$steps)
activity_all$interval <- as.numeric(activity_all$interval)
```

Now as the first step done, we will move to the next step. 
## 2: Histogram of the total number of steps taken each day
```{r}
# generate df with complete cases only
activity <- na.omit(activity_all)
activity$steps <- as.numeric(activity$steps)
activity$interval <- as.numeric(activity$interval)

stepsCnt <- aggregate(steps~date,data=activity,FUN="sum")
hist(stepsCnt$steps,col=1, main="Histogram of total number of steps per day", 
     xlab="Total number of steps in a day")
```

## 3: Mean and median number of steps taken each day

As Mean & Median are computed for all days below
```{r}
mean(stepsCnt$steps)
median(stepsCnt$steps)
```
And how about all the days? The mean of the steps by days would be interesting to see for all dates that are included in the dataset.
```{r}
meanSteps <- tapply(stepsCnt$steps, stepsCnt$date, mean)
medianSteps <- tapply(stepsCnt$steps, stepsCnt$date, median)
MeannMedian <- cbind(meanSteps, medianSteps)
MeannMedian
```

# 4. Time series plot of the average number of steps taken

What is the average daily activity pattern?

A time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
# aggregate steps as interval to get average number of steps in an interval across all days
table_interval <- aggregate(steps ~ interval, activity, mean)

# generate the line plot of the 5-minute interval (x-axis) and the average number of 
# steps taken, averaged across all days (y-axis)
plot(table_interval$interval, table_interval$steps, type='l', col=1, 
     main="Average number of steps averaged over all days", xlab="Interval", 
     ylab="Average number of steps")
```

# 5: The 5-minute interval that, on average, contains the maximum number of steps

```{r}
# find row id of maximum average number of steps in an interval
max_avg_steps_row <- which.max(table_interval$steps)

# get the interval with maximum average number of steps in an interval
table_interval [max_avg_steps_row, ]
```

# 6: Code to describe and show a strategy for imputing missing data
First find out how many NA values that we have in scope.
```{r}
# get rows with NA's
df_NA <- activity_all [!complete.cases(activity_all),]
# number of rows
nrow(df_NA)
```

For performing imputation, we replace the NA by the mean for that 5-minute interval. We already have this data in the data frame "table_interval".
We loop across the rows of the original data frame. If the steps value is NA for a row, we find the corresponding value of interval. We then look up the steps value from the other data frame "table_interval" for this value of interval and replace the NA value with it.

```{r}
table_interval <- aggregate(steps ~ interval, activity, mean)

# perform the imputation
for (i in 1:nrow(activity_all)){
  if (is.na(activity_all$steps[i])){
    interval_val <- activity_all$interval[i]
    row_id <- which(table_interval$interval == interval_val)
    steps_val <- table_interval$steps[row_id]
    activity_all$steps[i] <- steps_val
  }
}
```

Next - 

a. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
# aggregate steps as per date to get total number of steps in a day
table_imputed <- aggregate(steps~date,data=activity_all,FUN="sum")
```

## 7: Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
# create histogram of total number of steps in a day
hist(table_imputed$steps, col=1, main="(Imputed) Histogram of total number of steps per day", xlab="Total number of steps in a day")
```

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
mean(table_imputed$steps)
median(table_imputed$steps)
```

No change for Mean, but median is slightly different as we have more observations included.


## 8: Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends


First, we are creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

First convert date from string to Date class
```{r}
activity_all$date <- as.Date(activity_all$date, "%Y-%m-%d")
```
then add a new column indicating day of the week 
```{r}
activity_all$day <- weekdays(activity_all$date)
```

Add a new column called day type and initialize to weekday
If day is Saturday or Sunday, make day_type as weekend

```{r}
activity_all$day_type <- c("weekday")

for (i in 1:nrow(activity_all)){
  if (activity_all$day[i] == "Saturday" || activity_all$day[i] == "Sunday"){
    activity_all$day_type[i] <- "weekend"
  }
}

activity_all$day_type <- as.factor(activity_all$day_type)
```

Aggregate steps as interval to get average number of steps in an interval across all days

```{r}
table_imputed2 <- aggregate(steps ~ interval+day_type, activity_all, mean)
```

So the data is now ready. Time to plot.

```{r}
# make the panel plot for weekdays and weekends
library(ggplot2)
qplot(interval, steps, data=table_imputed2, geom=c("line"), xlab="Interval", 
      ylab="Number of steps", main="") + facet_wrap(~ day_type, ncol=1)
```

### Dont forget to cleanup the cache
```{r}
##Clearing the temp variables
rm(list=ls())
```
